Clinical Trial Design Using A Stopped Negative Binomial Distribution

### 2017-07-12 stat-and-prob-letters/snb-paper.tex
Lots of update.

Consider a prototypical early phase, single-arm clinical trial in which 
17 patients
are enrolled and treated. Suppose the Bernoulli probability of a patient 
responding to treatment is $p=0.2$ 
under the null hypothesis that the treatment is not effective.
If seven or more patients out of these 17 respond to the treatment then we 
reject this hypothesis and the treatment is deemed successful at 
a significance level of $0.1$.  If fewer than seven respond then the null 
hypothesis is not 
rejected and the treatment is deemed ineffective. The trial is 
realized as a sequence of Bernoulli($p$) samples stopping when either
the specified number of responders or non-responders is reached.

@@

\begin{figure}[bp!]
\centering
\includegraphics[width=\textwidth]{snb-first-plot.pdf}
\caption{
The distribution of the sample size in a trial that stops after seven patients
respond to treatment or 11 do not when $p=0.2$.
}
\label{fig:kane_viz}
\end{figure}

\begin{figure}[bp!]
\centering
\includegraphics[width=\textwidth]{mean-and-variance.pdf}
\caption{
The effect of varying $p$ between zero and one when the trial stops after
seven patients respond or 11 do not.
}
\label{fig:kane_viz}
\end{figure}

@@

The distribution of the number of trial enrollments is shown in 
Fig.~\ref{fig:kane_viz}. There is relatively little probability mass
for values of sample size equal to seven through 10 since $p$ is small and it 
is unlikely the treatment will succeed quickly.
%The probability mass is concentrated at the 
%$11^{\text{th}}$ and $12^{\text{th}}$ step since $p$ 
%is small and for most realizations reach 11 
%non-responders before reaching 2 responders.
Fig.~\ref{fig:kane_viz} shows the expected value and variance for the
number of trial enrollments varying $p$ between zero and one. When $p$ is
small then the treatment
is more likely to fail shortly after the $11^{\text{th}}$ enrollment.
When $p$ is large then the treatment is more likely to succeed and the 
number of enrollees approaches seven from above.

@@

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}.
\end{equation*}
The probability mass function is
\begin{equation} \label{eqn:pmf}
\mathbb{P} [Y=k] = S(k, p, s) \ I_{\{s \leq k \leq s+t-1\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \leq s+t-1 \}}
\end{equation}
where $I_{\{f\}}$ is the {\em indicator function}, taking the value 
of one if $f$ is true and zero otherwise, and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability mass. 
The CDF of the negative binomial is commonly expressed in terms of the
{\em regularized incomplete beta function} \citep{Olver2010}. Using standard
notation, (\ref{eqn:N}) is written as $\mathcal{I}_{1-p}(t, s)$.

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. In Fig.~\ref{fig:kane_viz} 
we consider a graphical illustration of the plot $X_k$ against
$k$. If the outcome of the $k$th patient responds to treatment then the process 
advances diagonally in the positive horizontal and vertical direction. 
If the $k$th patient does not respond
then the sample path advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$ corresponding to the
success and failure boundaries in Fig. \ref{fig:kane_viz}, respectively.

@@

To show (\ref{eqn:pmf}) sums to one, define
%\begin{align} \label{eqn:sum_proof}
%R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
%  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
%\end{align}
\begin{equation*} 
R = \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t).
\end{equation*}
If we substitute $i=k-s$ in the first summation and $j=k-t$ in the second then
$R$ can be written as the cumulative distribution function (CDF) of two
negative binomial distributions:
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

The regularized incomplete beta function satisfies
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Uppuluri1967}. Then
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof (\ref{eqn:pmf}) is the distribution of the
stopping time and is a valid probability mass function.
\end{proof}

@@

The SNB is a generalization of the negative 
binomial distribution. When $t$ is large then $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j\ |\ t \text{ is large\ }]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is moderate. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Peizer1968} and \cite{Best1974}. Examples are
shown in in Fig.~\ref{shapes.fig}.

@@

\begin{prop} \label{alpha_c}
Let $P_k$ be the posterior distribution of reaching the success endpoint
at the $k$th enrollment with density
\begin{equation*}
f_k(p) = \frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s-1}.
\end{equation*}
If $0 \leq \alpha_C \leq 1 - 1/k$ then 
\begin{equation}
\mathbb{P} \left[P_k \geq \alpha_C \right]  \leq
  \mathbb{P} \left[P_{k-1} \geq \alpha_C \right].
\end{equation}
\end{prop}
\begin{proof}
Consider the ratio of the posterior distributions
\begin{align*}
\frac{f_k(p)}{f_{k+1}(p)} &= 
  \frac{\frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s-1}}
    {\frac{ {k \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s}} \\
  &= \frac{ p^{\alpha +s -1} (1-p)^{k+\beta-s-1} }
    {k p^{\alpha +s -1} (1-p)^{k+\beta-s}} \\
  &= \frac{1}{k(1-p)}.
\end{align*}
This implies that the density $f_{k+1}(p)$ is greater than $f_k(p)$ for
all values of $0 \leq p \leq 1 - 1/k$. This implies the
integral of $f_{k+1}(p)$ must also be greater than $f_k(p)$ over a similar
interval. The result follows by realizing that this relationship is reversed
over the complement of the interval.
\end{proof}

@@

Keeping $s$ and $t$ constant for all time steps in the trial provides 
mathematical convenience since the distribution can be calculated in closed 
form. It should be noted though that any set of endpoints where every endpoint 
can be reached defines a valid probability distribution. This fact allows us 
to define new stopping criteria when defining success and failure endpoints.

One such set of stopping criteria is based on the confidence sets of the 
response rate from the previous section. Proposition (\ref{alpha_c}) shows 
$\mathbb{P_k} [P_k \geq \alpha_c] \leq \mathbb{P}[P_{k-1} \geq \alpha_c]$. 
That is, the probability the response rate is at least $p_{min}$ is increasing 
in $-k$. This observation implies we can guarantee that the response rate is 
greater than $p_min$ before reaching $s$ when a greater proportion of 
successful outcomes are observed.

@@

>>>>

Consider a prototypical early phase, single-arm clinical trial in which 
17 patients
are enrolled and treated. Suppose the Bernoulli probability of a patient 
responding to treatment is $p=0.2$ 
under the null hypothesis that the treatment is not effective.
If seven or more patients out of these 17 respond to the treatment then we 
reject this hypothesis and the treatment is deemed successful at 
a significance level of $0.1$.  If fewer than seven respond then the null 
hypothesis is not 
rejected and the treatment is deemed ineffective. The trial is 
realized as a sequence of indiependent Bernoulli($p$) samples stopping when 
either the specified number of responders or non-responders is reached.

@@

\begin{figure}[bp!]
\centering
\includegraphics[width=\textwidth]{snb-first-plot.pdf}
\caption{
The distribution of the sample size in a trial that stops after seven patients
respond to treatment or 11 do not when $p=0.2$.
}
\label{fig:kane_viz_1}
\end{figure}

\begin{figure}[bp!]
\centering
\includegraphics[width=\textwidth]{mean-and-variance.pdf}
\caption{
The effect of varying $p$ between zero and one when the trial stops after
seven patients respond or 11 do not.
}
\label{fig:kane_viz_2}
\end{figure}

@@

The distribution of the number of trial enrollments is shown in 
Fig.~\ref{fig:kane_viz_1}. There is relatively little probability mass
for values of sample size equal to seven through 10 since $p$ is small and it 
is unlikely the treatment will succeed quickly.
%The probability mass is concentrated at the 
%$11^{\text{th}}$ and $12^{\text{th}}$ step since $p$ 
%is small and for most realizations reach 11 
%non-responders before reaching 2 responders.
Fig.~\ref{fig:kane_viz_2} shows the expected value and variance for the
number of trial enrollments varying $p$ between zero and one. When $p$ is
small then the treatment
is more likely to fail shortly after the $11^{\text{th}}$ enrollment.
When $p$ is large then the treatment is more likely to succeed and the 
number of enrollees approaches seven from above.

@@

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}.
\end{equation*}
The probability mass function is
\begin{equation} \label{eqn:pmf}
\mathbb{P} [Y=k] = S(k, p, s) \ I_{\{s \leq k \leq s+t-1\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \leq s+t-1 \}}
\end{equation}
where $I_{\{f\}}$ is the {\em indicator function}, taking the value 
of one if $f$ is true and zero otherwise, and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability mass. 
The CDF of the negative binomial is commonly expressed in terms of the
{\em regularized incomplete beta function} \citep{Olver2010}. Using standard
notation, (\ref{eqn:N}) is written as $\mathcal{I}_{1-p}(t, s)$ and has
the property $\mathcal{I}_{1-p}(t, s) = 1 - \mathcal{I}_p(s, t)$
\citep{Uppuluri1967}.

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. In Fig.~\ref{fig:kane_viz_1} 
we consider a graphical illustration of the plot $X_k$ against
$k$. If the outcome of the $k$th patient responds to treatment then the process 
advances diagonally in the positive horizontal and vertical direction. 
If the $k$th patient does not respond
then the sample path advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$ corresponding to the
success and failure boundaries in Fig. \ref{fig:kane_viz_2}, respectively.

@@

To show (\ref{eqn:pmf}) sums to one, define
%\begin{align} \label{eqn:sum_proof}
%R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
%  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
%\end{align}
\begin{equation*} 
R = \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t).
\end{equation*}
Substitute $i=k-s$ in the first summation and $j=k-t$ in the second then
$R$ can be written as the cumulative distribution function (CDF) of two
negative binomial distributions:
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

Then
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof (\ref{eqn:pmf}) is the distribution of the
stopping time and is a valid probability mass function.
\end{proof}

@@

The SNB is a generalization of the negative 
binomial distribution. When $t$ is large then $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j ]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is moderate. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Peizer1968} and \cite{Best1974}. Examples are
shown in in Fig.~\ref{shapes.fig}.

@@

\begin{prop} \label{alpha_c}
Let $P_k$ be the posterior distribution of reaching the success endpoint
at the $k$th enrollment with density
\begin{equation*}
f_k(p) = \frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s-1}.
\end{equation*}
If $0 \leq \alpha_C \leq 1 - 1/k$ then 
\begin{equation}
\mathbb{P} \left[P_k \geq \alpha_C \right]  \leq
  \mathbb{P} \left[P_{k-1} \geq \alpha_C \right].
\end{equation}
\end{prop}
\begin{proof}
Consider the ratio of the posterior distributions
\begin{align*}
\frac{f_k(p)}{f_{k+1}(p)} &= 
  \frac{\frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s-1}}
    {\frac{ {k \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s}} \\
  &= \frac{ p^{\alpha +s -1} (1-p)^{k+\beta-s-1} }
    {k p^{\alpha +s -1} (1-p)^{k+\beta-s}} \\
  &= \frac{1}{k(1-p)}.
\end{align*}
This implies that the density $f_{k+1}(p)$ is greater than $f_k(p)$ for
all values of $0 \leq p \leq 1 - 1/k$. The
integral of $f_{k+1}(p)$ must also be greater than $f_k(p)$ over a similar
interval. The result follows by realizing that this relationship is reversed
over the complement of the interval.
\end{proof}

@@

Keeping $s$ and $t$ constant for all time steps in a trial provides 
mathematical convenience since the distribution can be calculated in closed 
form. It should be noted though that any set of endpoints where every endpoint 
can be reached by the process $\mathbf X$ defines a valid probability 
distribution. This fact allows us 
to define new stopping criteria when defining success and failure endpoints.

One such set of stopping criteria is based on the confidence sets of the 
response rate from the previous section. Proposition (\ref{alpha_c}) shows 
$\mathbb{P} [P_k \geq \alpha_c] \leq \mathbb{P}[P_{k-1} \geq \alpha_c]$. 
That is, the probability the response rate is at least $p_{min}$ is increasing 
in $-k$. This observation implies we can guarantee that the response rate is 
greater than $p_{min}$ before reaching $s$ when a greater proportion of 
successful outcomes are observed.

@@

As an example, consider the design of the hypothetical trial from the
first section. Assume that we are testing for a response rate of at least
0.2 and a uniform prior on the response rate then, by Proposition 
\ref{alpha_c}, the design implies a confidence of at least 98\% 
($\alpha_c = 0.02)$. That is, at each of the success endpoints, 
we believe the response rate is at least 0.2 with 98\% confidence. This also 
implies that $\beta = 2.24526 * 10^{-5}$.
Figure \ref{fig:bsprt} shows the success and failure boundaries for the 
SNB trial design from Section 1 along with the Bayesian SPRT analogue.

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{bsprt.pdf}
\caption{
Success and failure boundaries for the Bayesian SPRT analogue (BSPRT) and SNB
trial designs of the hypothetical trial.
}
\label{fig:bsprt}
\end{figure}
